Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2020-07-11 paper.tex
shrunk paper down to 16 pages again :)

%The Corona-virus warning app
%has taken a long time to develop being published only on
%16th June 2020. It
%was a quite costly project but this was mainly due to the management of
%Telekom and SAP being in the driving seat. But the app has been designed with great
The app was designed with great
attention on privacy: a distributed architecture \cite{cwa:arch} has been adopted after a long and
heated debate with supporters of a central architecture. The distributed architecture is
based on a very clever application design whereby users' phones broadcast
highly anonymized so called ``Ephemeral IDs'' at physical locations via the
Bluetooth Low Energy Beacon protocol.
The app saves those IDs of people in close proximity.
When at a later date an infected person reports his infection to a central server,
all clients are provided the means to reconstruct the IDs used by this particular
person's device over the last 14 days.
Therefore, any client can evaluate exposure risks locally and stored contact data has never to be shared.
%
%, the unique root ID is published
%and in the daily check all mobile phones connecting to the central server download
%the root IDs of infected people. Since the Ephemeral IDs can be mapped to the root ID
%all Ephemeral IDs that have been saved over the last 14 days allow users' phones to
%regularly check whether their user has been exposed to an infected person and issue
%a warning to the user and recommendation to contact health authorities.
%The warning issued by the Corona warning app entitles to
%having a Corona test done (which at the time of writing is not normally possible).

@@

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%The Corona-virus app has been produced based on a sophisticated security
%concept conceived by experts in the field %to our knowledge, no formal verification
%%
%and has strong claims with regard to mathematical support:
%
%``Such a design builds on strong, mathematically provable support for privacy and data protection goals [...]'' \TODOref{DP-3T Whitepaper p. 2} \smallskip\\
%  \NOTE{DP-3T on their design and how wisely chosen it is ;) We have to use this! Something along the l%ines of:}
%
However, there has been, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows to reveal weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or where and to what extend trade offs have to be made due to inherent vulnerabilities.
Given that the protocol implemented by the framework CWA runs on resembles only the most basic
of the three proposed by the DP-3T project,
a formal verified model that stresses the impact or limits of certain security measures
might help lend some more weight to appeals like \cite{dp3t:impl}
%to Apple and Google
to adopt the more sophisticated protocols.
%
%Thereby, we believe
%that our current work is useful to increase the trust in the Corona-virus app necessary for
%its wide adoption which in turn is crucial for it to be efficient.

@@

In this paper, we first provide some background in Section \ref{sec:background}:
we give a detailed overview of the development history %and security and privacy
and relevant parts of the Corona-virus warning app (Section \ref{sec:history})
We then introduce the Isabelle Infrastructure framework (Section \ref{sec:isainf}).
Next, we present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}). The found attack on the first abstract specification
motivates refinement. The formal definition of refinement for the Isabelle Infrastructure
framework is introduced and illustrated on the Corona-virus warning app (Section \ref{sec:ref})
before drawing some conclusions (Section \ref{sec:concl}).

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF) \cite{enf:proj}, jointly published by Apple and Google, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
Corona-virus warning app but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

There are, however, competing architectures noteworthy, namely protocols developed under the
roof of the \textit{Pan-European Privacy-Preserving Proximity Tracing} project (PEPP-PT) \cite{pepppt:github}, e.g.
PEPP-PT-ROBERT \cite{pepppt:ROBERT}.

Neither DP-3T nor PEPP-PT are synonymous for just one single protocol. Each project endorses
different protocols with unique properties in terms of privacy and data protection.
Yet, on a higher level of abstraction, it seems feasible to distinguish two basic architectures:
protocols as endorsed by PEPP-PT might be characterized as centralized architectures whereas
DP-3T-inspired protocols follow a (more) decentralized approach.\footnote{%
  DP-3T involves a central backend server. It is decentralized with regard to
  the collection and evaluation of contact information:
  In centralized architectures the server provides a risk scoring services, whereas decentralized
  approaches rely on local risk assessment and, thus, do not need to share contact information with
  the backend.}
\\

All DP-3T protocols roughly work the same:
%\subsubsection{Basic DP-3T protocol: How does it work?}
%  How does it relate to ENF/CWA? (broad strokes)}
%
Upon installation, the app generates secret daily seeds to derive \textit{Ephemeral Ids}
(EphIDs) from them. EphIDs are generated locally with cryptographic methods and cannot be connected
to one another but only reconstructed from the secret seed they were derived from.

@@

In the most simple (and insecure) protocol proposed by DP-3T %\cite{dp3t:wp}[p14ff]
this basically translates into publishing the daily seeds used to derive EphIDs from.
The protocol implemented by ENF and,
hence, CWA adopts this low-cost design \cite{cwa:arch}. DP-3T proposes two other, more sophisticated protocols that improve
privacy and data protection properties to different degrees but are more costly to implement.
%Aside from additional security measures,
%\TODOfn[styC]{In short: I want to name it but don't talk about it (since we're not really concerned with% this level of detail). But this sounds just sh...}
%like signing content the server provides as a general rule, the protocol implemented by ENF and,
%hence, CWA follows this low-cost design.\TODOref{CWA solution architecture ($\rightarrow$ github:CWA); probably ENF specification} DP-3T proposes two other, more sophisticated protocols that improve
%privacy and data protection properties to different degrees but are more costly.
%%TODOfn[styC]{``complicated'' sounds lame but ``costly'' is, perhaps, to simple/literal} to set up.

Figure \ref{fig:dp3tprot} illustrates the basic system architecture
along with some of the mitigation measures either implemented in CWA or proposed by DP-3T.

\begin{figure}[htb]
\vspace{-.5cm}
  \begin{center}
    \includegraphics[width=5in]{DP-3T_basic_protocol}
  \end{center}
\vspace{-.5cm}
  \caption{Decentralized privacy-preserving proximity tracing protocol of CWA}
  \label{fig:dp3tprot}
\vspace{-.5cm}
\end{figure}

@@

\section{Summary and discussion of relevance of the approach}
%{Conclusions}
\label{sec:concl}
%\subsection{Protection goals of attacks}
%\subsection
=======
\subsection{Protection goals of attacks}
Bluetooth based proximity tracing gives rise to a variety of noteworthy privacy and security issues that deserve formal consideration (and verification). The debate emerging between advocates of centralized architectures and those in favor of a decentralized approach in particular yields a lot of interesting material detailing different attacks and possible mitigation strategies: \cite{pepppt:dp3tana}, \cite{pepppt:dp3tresp}, \cite{dp3t:psre}.

In terms of attack scenarios, connecting an Ephemeral Id to a particular device is the first step into multiple different attacks, we want to classify as deanonymisation attacks:
\begin{itemize}
\item \textbf{Tracking a device}
  \cite{dp3t:psre}[p9], \cite{pepppt:dp3tana}[p8]\\
  Bluetooth enabled devices introduce some fundamental risks, that are not
  specific to proximity detection: If the Bluetooth signal is linked to a fixed
  identifier, say a MAC address, constant broadcasting makes
  a device trackable.
  Periodically changing Ephemeral Ids can be an effective mitigation\footnote{%
    In addition Bluetooth has to be configured to use rotating MAC addresses
    instead of fixed ones, as is the default in modern smartphone operating
    systems. Also, rotation of Bluetooth identifiers has to happen
    in sync with MAC address randomization; the Apple/Google proposal
    \cite{enf:bluetooth} ensures this.}
  as they limit the time frame in which a certain id can be used to
  track the corresponding device, hence, rendering the attack not impossible
  but significantly less severe.
\item \textbf{Identify infected individuals}
  \cite{dp3t:psre}[p5], \cite{pepppt:dp3tana}[p9]\\
  Like any notification about, e.\,g., a STD, might inform a spouse about the
  infidelity of the other, given the right circumstances, being notified about a
  possible exposure to COVID-19 might reveal illicit information to
  the recipient like who it was that tested positive for the virus.
  This is an inherent vulnerability of any such system,
  whether it is analogous or digital (digitally aided).\footnote{
    As DP-3T emphasizes, it it, indeed, inherent to proximity tracing in general
    \cite{dp3t:psre}[p6].
    The likelihood of said circumstances to occur, however,
    might increase when notifications are carried out automatically.
    Therefore, we feel that it is, nonetheless, worth considering.}
  In contrast to tracking a Bluetooth enabled device,
  this attack is not considerably mitigated by changing Ephemeral Ids:
  %Once connected to a device or individual, an Ephemeral Id might yield the
  %desired information for up to 14 days
  A positive test result can be reported for as much as 14 days after the
  compromised Ephemeral Id was used and still reveal the desired information.
\end{itemize}

\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=6in]{DP-3T_deanonymisation_attacks}
  \end{center}
  \caption{Two deanynomisation attacks: Tracking users and identifying confirmed COVID-19 cases}
  \label{fig:dp3tprot}
\end{figure}

\subsection{Summary and discussion of relevance of the approach}
We can establish in our formal framework an attack that even a system using changing
Ephemeral ids can be broken if the attacker physically follows a victim. This is a basic attack on
anonymity: a user's connection between his Ephemeral Ids and personal details (Iphone MAC or name)
is revealed to the attacker. The protection goal of privacy is thereby destroyed.

>>>>

%The Corona-virus warning app
%has taken a long time to develop being published only on
%16th June 2020. It
%was a quite costly project but this was mainly due to the management of
%Telekom and SAP being in the driving seat. But the app has been designed with great
The app was designed with great
attention on privacy: a distributed architecture \cite{cwa:arch} has been adopted %after a long and
%heated debate with supporters of a central architecture. The distributed architecture is
that is
based on a very clever application design whereby clients broadcast
highly anonymized so called ``Ephemeral IDs'' via Bluetooth and
store those ids of people in close proximity.
%highly anonymized so called ``Ephemeral IDs'' at physical locations via the
%Bluetooth Low Energy Beacon protocol.
%The app saves those ids of people in close proximity.
Infected persons report their infection by uploading their ids to a central server, providing all clients the means to assess exposure risk locally,
hence, stored contact data has never to be shared.
%Infected persons report their infection to a central server,
%which, in turn, provides all clients the means to reconstruct the ids
%used by reported cases over the last 14 days.
%
%When at a later date an infected person reports his infection to a central server,
%all clients are provided the means to reconstruct the ids used by this particular
%person's device over the last 14 days.
%Therefore, exposure risks can be evaluated locally and stored contact data has never to be shared.
%
%, the unique root ID is published
%and in the daily check all mobile phones connecting to the central server download
%the root IDs of infected people. Since the Ephemeral IDs can be mapped to the root ID
%all Ephemeral IDs that have been saved over the last 14 days allow users' phones to
%regularly check whether their user has been exposed to an infected person and issue
%a warning to the user and recommendation to contact health authorities.
%The warning issued by the Corona warning app entitles to
%having a Corona test done (which at the time of writing is not normally possible).

@@

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%The Corona-virus app has been produced based on a sophisticated security
%concept conceived by experts in the field %to our knowledge, no formal verification
%%
%and has strong claims with regard to mathematical support:
%
%``Such a design builds on strong, mathematically provable support for privacy and data protection goals [...]'' \TODOref{DP-3T Whitepaper p. 2} \smallskip\\
%  \NOTE{DP-3T on their design and how wisely chosen it is ;) We have to use this! Something along the l%ines of:}
%
However, there has been, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows to reveal weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or to what extend trade offs have to be made due to inherent vulnerabilities.
Given that the protocol implemented by CWA resembles only the most basic
of the three protocols proposed by the DP-3T project,
a formal verified model that stresses the impact or limits of certain security measures
might give more weight to appeals like \cite{dp3t:impl}
%to Apple and Google
to adopt the more sophisticated protocols.
%
%Thereby, we believe
%that our current work is useful to increase the trust in the Corona-virus app necessary for
%its wide adoption which in turn is crucial for it to be efficient.

@@

In this paper, we first provide some background in Section \ref{sec:background}:
we give a brief overview of related works
and the protocol of the Corona-virus warning app (Section \ref{sec:history}).
We then introduce the Isabelle Infrastructure framework (Section \ref{sec:isainf}).
Next, we present our model (Section \ref{sec:model}) and analysis of privacy and
attacks (Section \ref{sec:ana}). The found attack on the first abstract specification
motivates refinement. The formal definition of refinement for the Isabelle Infrastructure
framework is introduced and illustrated on the Corona-virus warning app (Section \ref{sec:ref})
before drawing some conclusions (Section \ref{sec:concl}).

@@

\section{Background and related work}
\label{sec:background}
\subsection{DP-3T and PEPP-PT}
\label{sec:history}
We are mainly concerned with the architecture and protocols proposed by the
DP-3T (\textit{Decentralized Privacy-Preserving Proximity Tracing}) project \cite{dp3t:github}.
The main reason to focus on this particular family of protocols is the
\textit{Exposure Notification Framework} (ENF), jointly published by Apple and Google \cite{enf:proj}, that adopts
%the main
core principles of the DP-3T proposal. This API is not only used in the German
Corona-virus warning app but has the potential of being widely adopted in future
app developments that might emerge due to the reach of players like Apple and Google.

There are, however, competing architectures noteworthy, namely protocols developed under the
roof of the \textit{Pan-European Privacy-Preserving Proximity Tracing} project (PEPP-PT) \cite{pepppt:github}, e.g.
PEPP-PT-ROBERT \cite{pepppt:ROBERT},
that might be characterized as centralized architectures.

Neither DP-3T nor PEPP-PT are synonymous for just one single protocol. Each project endorses
different protocols with unique properties in terms of privacy and data protection.
%Yet, on a higher level of abstraction, it seems feasible to distinguish two basic architectures:
%protocols as endorsed by PEPP-PT might be characterized as centralized architectures whereas
%DP-3T-inspired protocols follow a (more) decentralized approach.\footnote{%
%  DP-3T involves a central backend server. It is decentralized with regard to
%  the collection and evaluation of contact information:
%  In centralized architectures the server provides a risk scoring services, whereas decentralized
%  approaches rely on local risk assessment and, thus, do not need to share contact information with
%  the backend.}

\subsubsection{Basic DP-3T architecture}
%All DP-3T protocols roughly work the same:
%\subsubsection{Basic DP-3T protocol: How does it work?}
%  How does it relate to ENF/CWA? (broad strokes)}
%
Upon installation, the app generates secret daily seeds to derive \textit{Ephemeral Ids}
(EphIDs) from them. EphIDs are generated locally with cryptographic methods and cannot be connected
to one another but only reconstructed from the secret seed they were derived from.

@@

In the most simple (and insecure) protocol proposed by DP-3T %\cite{dp3t:wp}[p14ff]
this basically translates into publishing the daily seeds used to derive EphIDs from.
The protocol implemented by ENF and,
hence, CWA adopts this low-cost design \cite{cwa:arch}. DP-3T proposes two other, more sophisticated protocols that improve
privacy and data protection properties to different degrees but are more costly to implement.
%Aside from additional security measures,
%\TODOfn[styC]{In short: I want to name it but don't talk about it (since we're not really concerned with% this level of detail). But this sounds just sh...}
%like signing content the server provides as a general rule, the protocol implemented by ENF and,
%hence, CWA follows this low-cost design.\TODOref{CWA solution architecture ($\rightarrow$ github:CWA); probably ENF specification} DP-3T proposes two other, more sophisticated protocols that improve
%privacy and data protection properties to different degrees but are more costly.
%%TODOfn[styC]{``complicated'' sounds lame but ``costly'' is, perhaps, to simple/literal} to set up.
%\pagebreak

Figure \ref{fig:dp3tprot} illustrates the basic system architecture
along with some of the mitigation measures either implemented in CWA or proposed by DP-3T.
\begin{figure}
\vspace{-.5cm}
  \begin{center}
    \includegraphics[width=4in]{DP-3T_basic_protocol}
  \end{center}
\vspace{-.5cm}
  \caption{Decentralized privacy-preserving proximity tracing protocol of CWA}
  \label{fig:dp3tprot}
\vspace{-.5cm}
\end{figure}

\subsubsection{Attack scenarios}
There is a variety of noteworthy privacy and security issues. %that deserve formal consideration (and verification).
The debate among %emerging between
advocates of centralized architectures and those in favor of a decentralized approach in particular yields a lot of interesting material detailing different attacks and possible mitigation strategies: \cite{pepppt:dp3tana}, \cite{pepppt:dp3tresp}, \cite{dp3t:psre}.

In terms of attack scenarios, we focus on, what might be classified as deanonymisation attacks: Tracking a device (see \cite{dp3t:psre}[p9], \cite{pepppt:dp3tana}[p8]) and identifying infected individuals (\cite{dp3t:psre}[p5], \cite{pepppt:dp3tana}[p9]).
%As first step, both attacks start with deanonymizing Ephemeral Ids,
%i.\,e. linking them the particular devices.
Both attacks rely on deanonymisation of Ephemeral Ids,
i.\,e. linking ids to particular devices.

@@

\section{Summary and discussion of relevance of the approach}
%{Conclusions}
\label{sec:concl}
%\subsection{Protection goals of attacks}
We can establish in our formal framework an attack that even a system using changing
Ephemeral ids can be broken if the attacker physically follows a victim. This is a basic attack on
anonymity: a user's connection between his Ephemeral Ids and personal details (Iphone MAC or name)
is revealed to the attacker. The protection goal of privacy is thereby destroyed.
